---
title: Preliminary Process Mining Analysis Report
subtitle: Patient journey (Diabetes) - LNCC (Part 5/6)
author: "Marcelo Santos"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_notebook
--- 

```{r echo=FALSE}
library(dplyr)
library(bupaR)
library(bit64)
```

```{r echo=FALSE}
data_folder <- paste(getwd(), '/../datasets/', sep='')
# Opening dfevents (new dataset)
dfevents <- read.csv(paste(data_folder,"dfevents2015 - prenatal-parto-puerperio - 20181020.csv",sep=''), sep=',', header=T, stringsAsFactors = TRUE)
# Preparing fields to be used
dfevents$datacompleta <- as.POSIXct(strptime(as.character(dfevents$datacompleta), "%Y-%m-%d"))
dfevents$tipoestabelecimentoexecutante <- as.factor(dfevents$tipoestabelecimentoexecutante)
dfevents$nomeestabelecimentoexecutante <- as.factor(dfevents$nomeestabelecimentoexecutante)
dfevents$cepcompleto <- as.factor(dfevents$cepcompleto)
dfevents$nomeespecialidade <- as.factor(dfevents$nomeespecialidade)
dfevents$codigocid <- as.factor(dfevents$codigocid)
dfevents$nomecid <- as.factor(dfevents$nomecid)
dfevents$codigoprocedimento <- as.factor(dfevents$codigoprocedimento)
dfevents$nomeprocedimento <- as.factor(dfevents$nomeprocedimento)
dfevents$codigocboexecutante <- as.factor(dfevents$codigocboexecutante)
dfevents$nomecboexecutante <- as.factor(dfevents$nomecboexecutante)
dfevents$h2nomenivel2a <- as.factor(dfevents$h2nomenivel2a)
dfevents$codigocns <- as.factor(as.integer64(dfevents$codigocns))
```

Screening patients with diabetes-related complications
```{r}
# ICD codes: O240, O241, O242, O243, O244, O249
dfaux<-subset(dfevents,substring(codigocid,1,3)=='O24')
```

```{r echo=FALSE}
dfaux<-subset(dfaux, h2nomenivel2a=='CRS NORTE')
```

```{r echo=FALSE}
dfpacients<-group_by(dfaux, codigocns) %>% summarize(qtde=n())
dfevents<-subset(dfevents,codigocns %in% dfpacients$codigocns)
dfevents$codigocns <- as.factor(as.integer64(dfevents$codigocns))
c1<-nrow(dfpacients)
c2<-nrow(dfevents)
```

#### CRS NORTE Analysis

Number of CNS codes -> **`r c1`**    

Number of filtered records -> **`r c2`**



```{r echo=FALSE}
# Frequencia de codigocns e tipo de estabelecimento
dfcnstipoAll <- group_by(dfevents, h2nomenivel2a, codigocns, datacompleta, tipoestabelecimentoexecutante, nomecboexecutante) %>% summarize(qtde=n())
dfcnstipoAll$activity_id <- row.names(dfcnstipoAll)
dfcnstipoAll$status <- 1
dfcnstipoAll$nomecboexecutante <- substring(dfcnstipoAll$nomecboexecutante,1,35)
```

```{r echo=FALSE}
dfcnstipo <- dfcnstipoAll
```

```{r echo=FALSE}
# Criando um arquivo de log
logevents <-  eventlog(dfcnstipo,
    case_id = 'codigocns',
    activity_id = 'tipoestabelecimentoexecutante',
    activity_instance_id = 'activity_id',
    lifecycle_id = 'status',
    timestamp = 'datacompleta',
    resource_id = 'nomecboexecutante')
```
IMPORTANT: in order to achieve better process map visualizations we are applying a percentile cut off of 95% of existing trace coverage. It means that the process will look at the 95% of the paths (set of steps) most performed by citizens during their journey in the public healthcare system.  
```{r echo=FALSE}
# processando o diagrama de processo
a<- logevents %>%
    filter_trace_frequency(perc = 0.95)
    rm(logevents)
```
The Figure 17 illustrates the type of facilities and the absolute frequency of times people pass in this way during their journey in the public healthcare system in Sao Paulo municipality - CRS NORTE.

```{r echo=FALSE, fig.height=4, fig.width=9}
a %>% process_map()
```
Figure 17 - Process map of Prenatal, Childbirth and Postpartum activities related to Diabetes patients performed in the Municipality of São Paulo in 2015 - CRS NORTE.

The idle time during the paths followed by citizens is shown in the Figure 18.

```{r echo=FALSE, fig.height=4, fig.width=9}
# Frequency
a %>% process_map(type_nodes = frequency("absolute"),
            type_edges = performance(median, "days"))
```
Figure 18 - Process map including absolute frequency and idle time of Prenatal, Childbirth and Postpartum activities related to Diabetes patients performed in the Municipality of São Paulo in 2015 - CRS NORTE.  

Idle time during paths and relative frequency for the facilities type is illustrated in the Figure 19.

```{r echo=FALSE, fig.height=4, fig.width=9}
# Combinando frequencia relativa e média de dias
a %>% process_map(type_nodes = frequency("relative"),
            type_edges = performance(median, "days"))
```
Figure 19 - Process map including relative frequency and idle time registered during the Prenatal, Childbirth and Postpartum activities related to Diabetes patients performed in the Municipality of São Paulo in 2015 - CRS NORTE. 

The summary of antecedent and consequent type of facility is presented in the Figure 20.

```{r echo=FALSE, fig.height=6, fig.width=9} 
# análise de precedência
a %>%
  precedence_matrix(type = "relative_antecedent") %>% plot
```
Figure 20 - Summary of antecedent and consequent type of facility for the Prenatal, Childbirth and Postpartum activities related to Diabetes patients performed in the Municipality of São Paulo in 2015 - CRS NORTE.

```{r echo=FALSE}
rm(a, dfcnstipo, dfcnstipoAll, dfevents, dfaux, dfpacients)
```

### Comments


### References
(1) https://en.wikipedia.org/wiki/Process_mining
(2) https://www.qpr.com/blog/what-process-mining
(3) Van der Aalst WM. Process discovery: An introduction. InProcess Mining 2011 (pp. 125-156). Springer, Berlin, Heidelberg.

