---
title: Preliminary Process Mining Analysis Report
subtitle: Patient journey - LNCC (Part 2/6)
author: "Marcelo Santos"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_notebook
--- 

```{r echo=FALSE}
library(dplyr)
library(bupaR)
library(bit64)
```

```{r echo=FALSE}
data_folder <- paste(getwd(), '/../datasets/', sep='')
# Opening dfevents (new dataset)
dfevents <- read.csv(paste(data_folder,"dfevents2015 - prenatal-parto-puerperio - 20181020.csv",sep=''), sep=',', header=T, stringsAsFactors = TRUE)
# Preparing fields to be used
dfevents$datacompleta <- as.POSIXct(strptime(as.character(dfevents$datacompleta), "%Y-%m-%d"))
dfevents$tipoestabelecimentoexecutante <- as.factor(dfevents$tipoestabelecimentoexecutante)
dfevents$nomeestabelecimentoexecutante <- as.factor(dfevents$nomeestabelecimentoexecutante)
dfevents$cepcompleto <- as.factor(dfevents$cepcompleto)
dfevents$nomeespecialidade <- as.factor(dfevents$nomeespecialidade)
dfevents$codigocid <- as.factor(dfevents$codigocid)
dfevents$nomecid <- as.factor(dfevents$nomecid)
dfevents$codigoprocedimento <- as.factor(dfevents$codigoprocedimento)
dfevents$nomeprocedimento <- as.factor(dfevents$nomeprocedimento)
dfevents$codigocboexecutante <- as.factor(dfevents$codigocboexecutante)
dfevents$nomecboexecutante <- as.factor(dfevents$nomecboexecutante)
dfevents$h2nomenivel2a <- as.factor(dfevents$h2nomenivel2a)
dfevents$codigocns <- as.factor(as.integer64(dfevents$codigocns))
```

```{r echo=FALSE}
# Frequencia de codigocns e tipo de estabelecimento
dfcnstipoAll <- group_by(dfevents, h2nomenivel2a, codigocns, datacompleta, tipoestabelecimentoexecutante, nomecboexecutante) %>% summarize(qtde=n())
dfcnstipoAll$activity_id <- row.names(dfcnstipoAll)
dfcnstipoAll$status <- 1
dfcnstipoAll$nomecboexecutante <- substring(dfcnstipoAll$nomecboexecutante,1,35)
rm(dfevents)
```

#### CRS CENTRO OESTE Analysis

```{r echo=FALSE}
dfcnstipo <- subset(dfcnstipoAll, h2nomenivel2a=='CRS CENTRO OESTE')
```

```{r echo=FALSE}
# Criando um arquivo de log
logevents <-  eventlog(dfcnstipo,
    case_id = 'codigocns',
    activity_id = 'tipoestabelecimentoexecutante',
    activity_instance_id = 'activity_id',
    lifecycle_id = 'status',
    timestamp = 'datacompleta',
    resource_id = 'nomecboexecutante')
```

```{r echo=FALSE}
# processando o diagrama de processo
a<- logevents %>%
    filter_trace_frequency(perc = 0.95)
rm(logevents)
```

IMPORTANT: in order to achieve better process map visualizations we are applying a percentile cut off of 95% of existing trace coverage. It means that the process will look at the 95% of the paths (set of steps) most performed by citizens during their journey in the public healthcare system.

The Figure 5 illustrates the type of facilities and the absolute frequency of times people pass in this way during their journey in the public healthcare system in Sao Paulo municipality - CRS Centro Oeste.

```{r echo=FALSE, fig.height=4, fig.width=9}
a %>% process_map()
```
Figure 5 - Process map of Prenatal, Childbirth and Postpartum activities performed in the Municipality of São Paulo in 2015 - CRS CENTRO OESTE.

The idle time during the paths followed by citizens is shown in the Figure 6.

```{r echo=FALSE, fig.height=4, fig.width=9}
# Frequency
a %>% process_map(type_nodes = frequency("absolute"),
            type_edges = performance(median, "days"))
```
Figure 6 - Process map including absolute frequency and idle time of Prenatal, Childbirth and Postpartum activities performed in the Municipality of São Paulo in 2015 - CRS CENTRO OESTE.  

Idle time during paths and relative frequency for the facilities type is illustrated in the Figure 7.

```{r echo=FALSE, fig.height=4, fig.width=9}
# Combinando frequencia relativa e média de dias
a %>% process_map(type_nodes = frequency("relative"),
            type_edges = performance(median, "days"))
```
Figure 7 - Process map including relative frequency and idle time registered during the Prenatal, Childbirth and Postpartum activities performed in the Municipality of São Paulo in 2015 - CRS CENTRO OESTE. 

The summary of antecedent and consequent type of facility is presented in the Figure 8.

```{r echo=FALSE, fig.height=6, fig.width=9} 
# análise de precedência
a %>%
  precedence_matrix(type = "relative_antecedent") %>% plot
```
Figure 8 - Summary of antecedent and consequent type of facility for the Prenatal, Childbirth and Postpartum activities performed in the Municipality of São Paulo in 2015 - CRS CENTRO OESTE.

```{r echo=FALSE}
rm(a, dfcnstipo, dfcnstipoAll)
```

### Comments


### References
(1) https://en.wikipedia.org/wiki/Process_mining
(2) https://www.qpr.com/blog/what-process-mining
(3) Van der Aalst WM. Process discovery: An introduction. InProcess Mining 2011 (pp. 125-156). Springer, Berlin, Heidelberg.

